---
title: "Exploring Forest Fire Data"
author: "Abe Ceasar Perez"
date: "March 21, 2022"
output: html_document
---

### Understanding Forest Fire Data in Portugal

This project aims to identify possible factors which lead to the occurrence of forest fires in Portugal. We'll be working as a researcher to supplement additional findings in a study conducted by P. Cortez which is linked here: <http://www3.dsi.uminho.pt/pcortez/fires.pdf>

In addition to the objectives above, we'll be applying the following concepts in this project:
- Barplots
- Scatterplots
- Multiple charting

### Initial Findings

Based on the initial results of this project, forest fires tend to occur in August and September, which usually happens on weekends (Saturday, Sunday) or nearing weekends (Friday). Factors such as DC, DMC, etc. were found to have an influence in the occurrence of forest fires for the given months. 

Apart from time, the area of fire was also investigated to determine any correlation against other variables. Factors such as DC, DMC, FFMC, Relative Humidity, and Wind were found to have an influence in the spread of a forest fire.

The process of this research is outlined below:

### Loading the Data

The forest fire data was obtained in the link below -  <https://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/>. This data was obtained from the author of the researcher, P. Cortez in his research of predicting forest fires.

We'll start off by loading the needed packages for this project followed by the actual data obtained from the website above. Initial inspection shows that the dataset has around 517 rows and 13 columns/variables.

```{r load_data, echo=TRUE, warning=FALSE}

setwd("C:\\Users\\Abe\\Desktop\\Dataquest modules\\Data Analyst in R\\Guided Projects\\project_4")

pacman::p_load("purrr", "tidyr","readr","dplyr","ggplot2","lubridate", "stringr") #loading all packages

forest_fires <- read_csv("forestfires.csv", show_col_types = FALSE) #loading the data w/o messages
dim(forest_fires) #get the dimensions of the dataframe
```

### Inspecting the Columns

To better understand the dataset as a whole, we also included the definitions for each variable after a brief research. Here are the definitions of each data below:

- X: X-axis spatial coordinate within the Montesinho park map: 1 to 9
- Y: Y-axis spatial coordinate within the Montesinho park map: 2 to 9
- month: Month of the year: 'jan' to 'dec'
- day: Day of the week: 'mon' to 'sun'
- FFMC: Fine Fuel Moisture Code index from the FWI system: 18.7 to 96.20
- DMC: Duff Moisture Code index from the FWI system: 1.1 to 291.3
- DC: Drought Code index from the FWI system: 7.9 to 860.6
- ISI: Initial Spread Index from the FWI system: 0.0 to 56.10
- temp: Temperature in Celsius degrees: 2.2 to 33.30
- RH: Relative humidity in percentage: 15.0 to 100
- wind: Wind speed in km/h: 0.40 to 9.40
- rain: Outside rain in mm/m2 : 0.0 to 6.4
- area: The burned area of the forest (in ha): 0.00 to 1090.84

```{r inspect_cols, echo=TRUE}

columns <- colnames(forest_fires) #store column names in another variable
columns

```

### Inspecting the Rows

After inspecting each columns in the dataset, we'll move on to inspecting the rows of the dataset. 
Each row in this dataset represents an occurrence of fire on a certain location and a certain date
including the measurements associated with that incident.

```{r inspect_rows, echo=TRUE}

head(forest_fires, 10) #prints out the first 10 rows of the dataset

```

FFMC, DMC, and DC all relate to the possible sources of forest fires.
Most of the sources pertained in these metrics are from wet to dry
sources, with dry sources being the most hazardous sources

```{r inspect_rows1, echo=TRUE}

head(forest_fires[,c(5:7)], 10) #prints out the first 10 rows of FFMC, DMC, and DC

```

Temp, RH, wind, and rain can be considered factors which may
either contribute or mitigate the existence of a given fire

```{r inspect_rows2, echo=TRUE}

head(forest_fires[,c(9:12)], 10) #prints out the first 10 rows of temp, RH, wind, and rain

```

Lastly, area and ISI pertain to the current existence of forest fire,
how it is currently being spread, and where it currently is being mitigated.

```{r inspect_rows3, echo=TRUE}

head(forest_fires[, c(8,13)], 10) #prints out the first 10 rows of area and ISI

```

### Modifying the Date Columns

Since we'll be analyzing data based on dates, we'll also be looking into the date columns to see if the data is apppropriate for the analysis. Looking into the date columns, we can see that the months and days are arranged in an alphabetical order and not in the conventional order of dates.

```{r inspect_dates, echo=TRUE}

#Pulls unique month/weekday values
forest_fires %>% pull(month) %>% unique
forest_fires %>% pull(day) %>% unique

```

For that case, we'll be modfiying these variables by turning these columns into ordered factors, i.e. variables which take into account the assigned order by the user. 

We'll start first by creating the factor labels for months and weekdays arranged in the correct order:

```{r modifying_dates, echo=TRUE}

factored_months <- str_to_lower(month(1:12, label=TRUE)) #gets all distinct month names in calendar order and changes the months to lowercase to mimic the month vector
factored_months

dates <-  unlist(map(1:7, function(x) as.character(ymd("2022-03-21") + 5 + x))) #gets all dates weekdays based on the reference date (March 27, Sunday to April 2, Saturday)
factored_days <- str_to_lower(wday(dates, label=TRUE, abbr=TRUE)) #changes the dates above into weekday names, similar to the data from Forest Fires
factored_days
```

After creating the factor labels, we'll then apply these changes to the dataset in preparation for our analysis.

```{r clean_data, echo=TRUE}

forest_fires <- forest_fires %>% mutate(month = factor(month, levels = factored_months), day = factor(day, levels = factored_days)) #changes month/day columns into ordered factors

#double-checking to make sure changes have been applies
forest_fires %>% pull(month) %>% unique
forest_fires %>% pull(day) %>% unique

```

### Which months do forest fires happen the most?

After the data has been cleaned, we'll now proceed with visualizing the data first to identify which months forest fires occur the most. We'll start first with creating a monthly table which contains the total number of forest fires for each month.

```{r table_monthly_data, echo=TRUE}

fires_per_month <- forest_fires %>% group_by(month) %>% summarize(Count = n()) #creates a summary table for forest fires per month
fires_per_month

```

We'll then proceed with plotting the data using the barplot in order to answer our inquiry above. We'll be indicating the months of interest below using a darker color. Based on the initial analysis, August and September seem to be the hotspot for forest fires.

```{r plot_monthly_data, echo=TRUE}

fires_per_month %>% ggplot(aes(x=month, y=Count)) + 
      geom_col(fill="burlywood") + #creates a barplot
      geom_col(data=subset(fires_per_month, month=="aug" | month == "sep"), fill="coral3") +        #highlights the months with the most forest fires
      labs(title = "Which months do forest fires happen the most?", x = "Month", y="No. of fires") +         theme_classic() + 
      theme(plot.title = element_text(face="bold"))
```

### Which days of the week do forest fires happen the most?

We'll now proceed with visualizing the data based on weekdays to identify which day of the week forest fires usually occur. Similar to the step above, we'll be creating a daily table which contains the total number of forest fires that occured on a given day.

```{r table_daily_data, echo=TRUE}

fires_per_day <- forest_fires %>% group_by(day) %>% summarize(Count = n()) #creates a summary table for forest fires per day
fires_per_day

```

We'll then proceed with plotting the data using the barplot in order to answer our inquiry above. We'll be indicating the months of interest below using a darker color. Based on the initial analysis, Fridays, Saturdays, and Sundays seem to be the most common days where forest fires had occurred.

```{r plot_daily_data, echo=TRUE}

fires_per_day %>% ggplot(aes(x=day, y=Count)) + 
      geom_col(fill="burlywood") + 
      geom_col(data=subset(fires_per_day, day=="sun" | day == "fri" | day == "sat"), fill="coral3") +        #highlights days with the most forest fires
      labs(title = "Which days of the week do forest fires happen the most?", 
           x = "Day", y="No. of fires") + 
      theme_classic() + 
      theme(plot.title = element_text(face="bold"))

```

### Which among the Factors contributed to the August/September Forest Fires?

Going back to the monthly analysis, we'll be exploring other variables which may have contributed to the increase in forest fires for the aforementioned months. We'll be considering metrics such as FFMC, DMC, DC, etc. to see if there is any correlations with the occurrence of forest fires. 

We'll start off by summarizing the monthly data along with the variables of interest using averages:

```{r table_monthly_vars, echo=TRUE}

fires_per_month_pivot <- pivot_longer(data = forest_fires[,c(3,5:12)], 
                                      cols = colnames(forest_fires)[5:12], 
                                      names_to = "variables", 
                                      values_to = "values") #combines all variables of interest into one column as well as their respective values

fires_per_month_pivot_sum <- fires_per_month_pivot %>% group_by(month, variables) %>% summarize(values = round(mean(values, na.rm=TRUE),2)) #aggregates all metrics using averaging

fires_per_month_pivot_sum

```

The data is then plotted using a bar chart for each variable to see any evident trends. Findings are summarized as below:

```{r plot_monthly_vars, echo=TRUE, fig.width=12}

fires_per_month_pivot_sum %>% ggplot(aes(x=month, y=values)) + 
      geom_col(fill="burlywood") + 
      geom_col(data=subset(fires_per_month_pivot_sum, month=="aug" | month == "sep"), fill="coral3") +
      #highlights the months of interest
      facet_wrap(vars(variables), scales = "free_y", nrow=2) + #y-axis is scaled based on the data for each variable
      labs(title = "Average Forest Fire Incidence Metrics per Month", x = "Month", y="Values") +              theme_classic() + 
      theme(plot.title = element_text(face="bold")) +
      theme_bw() + 
      theme(plot.title = element_text(face="bold")) 

```

- Based on the graph above, the variables DC, DMC, and FMC were found to be somehow
  correlated with the August and September forest fires, meaning, it's more likely that fuel sources were found to be the driest on those months  which could've triggered the fires. 

- Given the amount of dry materials present, this could've also triggered high ISI, meaning, more areas   for fires to spread.

- Rain did not seem to occur for the majority of the period except for August. Despite this, August still had the most forest fires for the said month.

- Relative humidity and wind were found to be the lowest in this periods, thus, supporting providing more evidence of drought in these periods
  
- Lastly, temperature is also observed to be near the peak of those months which led to the high occurrence of forest fires in these periods.

### Which Variables correlate to the Area of Forest Fire Spread?

We'll now move to the last part of the analysis which is determining if there are any correlations between area and the forest fire variables. We'll start first by creating another dataframe which includes the values for area and the columns including all the variable names and the values associated with them:

```{r table_monthly_corr, echo=TRUE}

fires_per_month_corr <- pivot_longer(data = forest_fires[c(13, 5:12)], 
                                     cols =  colnames(forest_fires)[5:12], 
                                     names_to = "variables", 
                                     values_to = "values") #filters area and other variables and pivots the variables into one column

fires_per_month_corr

```

After creating the dataframes, we'll now generate the initial visualizations using scatterplots and linear trends to see if there are any evident relationships. Based on the initial findings below, there is no definite trend to be observed for each of the variables. This is mainly due to the presence of outliers/extreme data points or in this context, forests with high area values which may have affected the visualization for each factor.

```{r plot_monthly_data0, echo=TRUE, fig.width=12}

fires_per_month_corr %>% ggplot(aes(x=values, y=area)) + 
      geom_point(alpha=0.5) + 
      geom_smooth(method="lm", se=FALSE) + #fits a linear trend above the scatterplot for each variable
      facet_wrap(vars(variables), scales="free_x", nrow=2) + #plots the data by variable and arranges them into two rows
      labs(title = "Relationship bet. Variables and Area of Forest Fire", x = "Values", y="Area") +              theme_classic() + 
      theme(plot.title = element_text(face="bold")) +
      theme_bw() + 
      theme(plot.title = element_text(face="bold"))

```

### Re-calibrating the Analysis

Since the initial graph do not present any significant findings, we'll then proceeed with removing certain data points to tune the analysis.

We'll be removing these outliers, i.e. forests with areas greater than 100 HA and re-conduct the analysis. Based on the results below, there seems to be a slight trend in FFMC, temperature, and wind against area of fire.


```{r plot_monthly_data1, echo=TRUE, fig.width=12}

#Plots only the areas with at most 100 HA
fires_per_month_corr %>% filter(area <= 100) %>% ggplot(aes(x=values, y=area)) + 
  geom_point(alpha=0.5) + 
  geom_smooth(method="lm", se=FALSE) + 
  facet_wrap(vars(variables), scales="free_x", nrow=2) + 
  labs(title = "Relationship bet. Variables and Area of Fire", x = "Values", y="Area") +                  theme_classic() + 
  theme(plot.title = element_text(face="bold")) +
  theme_bw() + 
  theme(plot.title = element_text(face="bold"))

```

We'll also consider removing data points equal to 0 i.e. forests with areas equal to 0 HA and re-conduct the analysis. Based on the results below, the trend is still difficult to observe, though there is still a slight increase in trend based on FMC and temperature.


```{r plot_monthly_data2, echo=TRUE, fig.width=12}

#Plots only the areas greater than 0 HA
fires_per_month_corr %>% filter(area > 0) %>% ggplot(aes(x=values, y=area)) + 
    geom_point(alpha=0.5) +      
    geom_smooth(method="lm", se=FALSE) + 
    facet_wrap(vars(variables), scales="free_x", nrow=2) +
    labs(title = "Relationship bet. Variables and Area of Fire", x = "Values", y="Area") +                  theme_classic() + 
    theme(plot.title = element_text(face="bold")) +
    theme_bw() + 
    theme(plot.title = element_text(face="bold"))

```

Lastly, we'll consider areas for a specified range. Based on the previous analyses, it would be ideal to select areas between 20 HA to 100 HA since these data removes both the outliers and the most frequent data points in the dataset. Based on the results below, relationships are now more evident in this graph. For areas bet. 20 to 100, there is an increasing trend in DMC, FFMC, and temperatures while there is a decreasing trend in relative humidity and wind.

```{r plot_monthly_data3, echo=TRUE, fig.width=12}

#Plots only the areas with HA ranging from 20 to 100
fires_per_month_corr %>% filter(area <= 100 & area >= 20) %>% ggplot(aes(x=values, y=area)) +                geom_point(alpha=0.5) + 
      geom_smooth(method="lm", se=FALSE) + 
      facet_wrap(vars(variables), scales="free_x", nrow=2) +
      labs(title = "Relationship bet. Variables and Area of Fire", x = "Values", y="Area") +                  theme_classic() + 
      theme(plot.title = element_text(face="bold")) +
      theme_bw() + 
      theme(plot.title = element_text(face="bold"))

```

### Conclusion

The main goal of this project is to identify possible factors which lead to the occurrence of forest fires in Portugal. In this project we considered all the possible indicators based on previous research in order to determine if there were any possible realationships that can be visualized in the analysis.

Based on the analysis above, Q3 months such as August and September became the hotspot months for forest fires. These months were considered the dry months of the year since various indicators such as DC, DMC, and FMC (which pertains to the dryness of the hazardous materials) were found to be the highest for these months. Other factors also point to the weather such as the temperature, wind, and humidity which all support the earlier hypothesis.

We've also looked into the relationship between the area of fire and the forest fire indicators and based on the analyses above, factors such as DMC, FFMC, and temperature are more likely to propagate the spread of forest fires while humidity and wind are confirmed to mitigate the spread of the forest fires (assuming that the area of spread is from 20 to 100 HA).

Given these findings, we recommend to include these into the analysis in order to better understand visually how certain factors can affect the presence of forest fires.
<br>
<br>
<br>



